### 2015-11-02 thesis-proposal/problem.tex
Tightened text of proposal intro, contributions sections

%Well-designed general-purpose programming languages give library providers the power to  modularly express a wide variety of useful constructs using a comparatively small set of primitives. %New primitives, and thus new dialects of these languages, should arise very rarely. 
%New primitives (and new \emph{dialects} of the language). %Extending a general-pandurpose language with new primitives, i.e.  forming an extended \emph{dialect} of the language, should, given a sufficiently expressive such language, be exceedingly rare. %Indeed, a diversity of dialects around a language can be taken as evidence against claims of  generality.%, this can be taken as evidence against claims of its generality. %Not all useful abstractions can be  realized, when considered comprehensively, as libraries.% The problem of achieving both simplicity and generality simultaneously with a single system remains unsolved. 
%Many languages claim to achieve this ideal. 
%For example, 
%If we measure the generality of a language by how frequently  dialects of the language emerge, 
%all major contemporary languages seem to leave room for improvement. 
%General-purpose programming languages are a dime a. 
%Among the chief goals in the design of a general-purpose programming languages must achieve \emph{stability}: they  very rarely needs to be extended or forked into dialects by the user community surrounding it. Instead, its users are able to modularly  express a wide variety of useful constructs in libraries, using the comparatively small set of primitives that the language builds in. 
Functional programming language designers have long studied small typed lambda calculi to develop a principled understanding of language-theoretic issues, like type safety, and to examine the mathematical character of language primitives of interest in isolation. These studies have then informed the design of ``full-scale''\footnote{Throughout this work, words and phrases that should be read as having an intuitive or informal meaning, rather than a strict mathematical meaning, will be introduced with quotation marks.} functional languages, which combine several such primitives and also introduce various generalizations and primitive ``embellishments''  motivated by a consideration of human factors. 
For example, major functional languages like Standard ML (SML) \cite{mthm97-for-dart,harper1997programming}, OCaml \cite{ocaml-manual} and Haskell \cite{jones2003haskell} all primitively build in record types, generalizing the nullary and binary product types that suffice in simpler calculi, because explicitly labeled components are cognitively useful to human programmers. Similarly, these languages build in derived syntactic forms (colloquially, ``syntactic sugar'') that decrease the syntactic cost of working with common library constructs, like lists.

The hope amongst many language designers is that a limited number of primitives like these will suffice to produce a ``general-purpose'' programming language, i.e. one where programmers can direct their efforts almost exclusively toward the development of libraries for a wide variety of application domains. However, a stable language design that fully achieves this ideal has yet to emerge, as evidenced by the diverse array of ``dialect'' languages that continue to proliferate around all major contemporary languages. 
In fact, tools that aid in the construction of so-called  ``domain-specific'' language dialects (DSLs)\footnote{In some parts of the literature, such dialects are called ``external DSLs'', to distinguish them from  ``internal'' or ``embedded DSLs'', which are actually  library interfaces that only ``resemble'' distinct dialects \cite{fowler2010domain}.} seem only to be becoming increasingly prominent. 
%\subsection{Why are there so many dialects?}
%{This calls for an investigation}: why is it that programmers and researchers are still so often unable to satisfyingly express the constructs that they need in libraries, as modes of use of the ``general-purpose'' primitives already available in major languages today, and instead see a need for new language dialects?

Although some of these dialects extend the type structure of the language that they are based on, the most common situation, and the one that will be our focus in this work, is that the existing semantic primitives suffice and it is only the \emph{syntactic cost} of expressing one or more constructs of interest using these primitives  that is too high. In response, library providers construct \emph{syntactic dialects} -- dialects that introduce only new derived syntactic forms. 
%Put another way, syntactic dialects can be specified by a context-independent expansion to the existing language that they are based on. 
For example, Ur/Web is a syntactic dialect of Ur (a language that itself descends from ML \cite{conf/pldi/Chlipala10}) that builds in derived forms for SQL queries, HTML elements and other datatypes used in the domain of web programming \cite{conf/popl/Chlipala15}. %Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined. 
This is not an isolated example -- there are many other types of data that could similarly benefit from the availability of specialized derived forms (we will consider another example, regular expression patterns, in Sec. \ref{sec:syntax-examples}). 
Tools like Camlp4 \cite{ocaml-manual}, Sugar* \cite{erdweg2011sugarj,erdweg2013framework} and Racket \cite{Flatt:2012:CLR:2063176.2063195}, which we will discuss in Sec. \ref{sec:syntax-existing-approaches}, have lowered the engineering costs of constructing syntactic dialects in such situations, further contributing to their proliferation.

%More advanced dialects introduce new type structure, going beyond what is possible with only new derived forms. As a simple example, the static and dynamic semantics of records cannot be expressed by context-independent expansion to a language with only nullary and binary products. Various languages have explored ``record-like'' primitives that go further, supporting functional update operators, width and depth coercions (sometimes implicit)%\cite{Cardelli:1984:SMI:1096.1098}
%, methods, prototypic dispatch and other such ``semantic embellishments'' that in turn cannot be expressed by context-independent expansion to a language with only standard record types (we will detail an  example in Sec. \ref{sec:metamodules-motivating-examples}). OCaml primitively builds in the type structure of polymorphic variants, open datatypes and  operations that use format strings like $\mathtt{sprintf}$ \cite{ocaml-manual}. ReactiveML builds in primitives for functional reactive programming \cite{mandel2005reactiveml}. ML5 builds in high-level primitives for distributed programming based on a modal lambda calculus \cite{Murphy:2007:TDP:1793574.1793585}. Manticore \cite{conf/popl/FluetRRSX07} and AliceML  \cite{AliceLookingGlass} build in parallel programming primitives with a more elaborate type structure than is found in simpler accounts of parallelism. 
%MLj builds in the type structure of the Java object system (motivated by a desire to interface safely and naturally with Java libraries) \cite{Benton:1999:IWW:317636.317791}. Other dialects do the same for other foreign languages, e.g. Furr and Foster describe a dialect of OCaml that builds in the type structure of C \cite{Furr:2005:CTS:1065010.1065019}. Tools like proof assistants and logical frameworks are used to specify and reason metatheoretically about dialects like these, and tools like compiler generators and language frameworks \cite{erdweg2013state} lower their implementation cost, again contributing to their proliferation.

\subsection{Dialects Considered Harmful}
% express record types as syntactic sugar over the simply-typed lambda calculus with  binary product types.\footnote{Pairs can of course be expressed as syntactic sugar atop records, though one could argue that using binary products as the more primitive concept is simpler.} The static semantics need to be extended with new type and term operators. However, the simplest way to express the dynamic semantics of the newly introduced term operators is by translation to nested binary products, so we can leave the operational semantics alone. \todo{fill this out} %For example, there are dozens of constructs that go by the name of ``records'' in various languages, each defined by a slightly different collection of primitive operations. \todo{examples} %, encouraged  historically  by the availability of tools like compiler generators and,  more recently, language workbenches \cite{workbenches} and DSL frameworks \cite{dsl}. Unfortunately, taking this approach makes it substantially more difficult for clients to import high-level abstractions orthogonally. 
% test 
Some  view the ongoing proliferation of dialects described above as harmless or even as desirable, arguing that programmers can simply choose the right dialect for the job at hand \cite{journals/stp/Ward94}. However, this ``dialect-oriented'' approach is, in an important sense, anti-modular: a library written in one dialect cannot, in general, safely and idiomatically interface with a library written using another dialect. 
Addressing this interoperability problem requires somehow ``combining'' the  dialects into a single language \cite{Benton:1999:IWW:317636.317791}. However, in the most general setting where the dialects in question might be specified by judgements of arbitrary form, this is not a well-defined notion. Even if we restrict our interest to dialects specified using formalisms that do operationalize some notion of dialect combination, there is generally no guarantee that the combined dialect will conserve important metatheoretic properties that can be established about the dialects in isolation. %In other words, any putative ``combined language'' must formally be considered a  distinct system for which one must derive essentially all metatheorems of interest anew, guided only informally by those derived for the dialects individually. %There is no well-defined mechanism for constructing such a ``combined language'' in general. 
For example, consider two syntactic dialects, one specifying derived syntax for finite mappings, the other specifying a similar syntax for \emph{ordered} finite mappings. Though each dialect can be specified by an unambiguous grammar in isolation, when these grammars are na\"ively  combined by, for example, Camlp4,  ambiguities  arise. 
%It is thus infeasible to simply allow different contributors to a software system to choose their own favorite dialect for each component they are responsible for. 
%It it clear that dialects are better rhetorical devices than practical engineering artifacts. 
Due to this paucity of modular reasoning principles, the ``dialect-oriented'' approach is problematic for software development ``in the large''. %Large software projects and software ecosystems must pick a single language that does provide powerful modular reasoning principles and, to benefit from them, stay inside it.

Dialect designers have instead had to take a less direct approach to have an impact on large-scale software development: they have had to convince the designers in control of comparatively popular languages, like OCaml and Scala, to include some suitable variant of the primitives they've espoused into backwards compatible language revisions. %These decisions are increasingly influenced by community processes, e.g. the Scala Improvement Process.  %This approach concentrates power as well as responsibility over maintaining metatheoretic guarantees in the hands of a small group of language designers, though increasingly influenced by various community processes (e.g. the Scala Improvement Process). 
%Dialects thus serve the role of rhetorical vehicles for new ideas, rather than direct artifacts. 
%Over time, accepting such extensions has caused these languages to balloon in size. 
This \emph{ad hoc} approach is not sustainable, for three main reasons. First,  there are simply too  many potentially useful such primitives, and many of these are only relevant in relatively narrow application domains (for derived syntax, our group has  gathered initial data speaking to this \cite{TSLs}). Second, primitives introduced earlier in a language's lifespan can end up monopolizing finite ``syntactic resources'', forcing subsequent primitives to use ever more esoteric forms. And third, primitives that prove after some time to be flawed in some way cannot be removed or replaced without breaking backwards compatibility. %Because there is no data about how useful a construct is in practice until it is available in a major language, decisions about which constructs to include are often informed only by intuition.
%Recalling the words of  Reynolds, which are clearly as relevant today as they were almost half a century ago \cite{Reynolds70}: %This approach is antithetical to the ideal of a truly \emph{general-purpose language} described at the beginning of this section.
%\newpage
%\begin{quote}\textit{The recent development of programming languages suggests that the simul\-taneous achievement of simplicity 
%and generality in language design is a serious unsolved 
%problem.}\begin{flushright}--- John Reynolds (1970)\end{flushright}
%\end{quote}

This suggests that language designers should strive to keep general-purpose languages small, stable and free of \emph{ad hoc} primitives. This leaves  two possible paths forward. One, exemplified (arguably) by SML, is to simply eschew further ``embellishments'' and settle on the existing primitives, which might be considered to sit at a ``sweet spot'' in the overall language design space. % (accepting that in some circumstances, this trades away expressive power or leads to  high syntactic cost). 
The other path forward is to search for a small number of highly general primitives that allow us degrade many of the constructs that are built primitively into dialects today instead to modularly composable library constructs. 
Encouragingly, primitives of this sort do occasionally arise. For example, a recent revision of OCaml added support for  generalized algebraic data types (GADTs), based on research on guarded recursive datatype constructors \cite{XiCheChe03}. Using GADTs, OCaml was able to move some of the \emph{ad hoc} machinery for typechecking operations that use format strings, like \texttt{sprintf}, out of the language and into a library. However, syntactic machinery remains  built in.

%Similarly, it recently introduced ``open datatypes'', which subsume its previous more specialized exception type, and captures many use cases for .

%Viewed ``dually'', one might equivalently ask for a language that builds in a core that is as small as possible, but provides expressive power comparable to languages with much larger cores. This is our goal in the work being proposed.

%\vspace{-10px}
\section{Proposed Contributions}
%Our broad aim in the work being proposed is to introduce primitive language mechanisms that give library providers the ability to  express new syntactic expansions as well as new types and operators in a safe and modularly composable manner. 
Our aim in the work being proposed is to introduce primitive language constructs that take further steps down the second path just described by reducing the need for syntactic dialects. In particular, we plan to introduce the following primitives:
% By supporting the primitives that we introduce, 1) Verse will be smaller than comparable languages like ML and Scala, and 2) dialect formation will be less frequently necessary. In other words, these primitives reduce the need for many others:%This eliminates the needs to build in fewer \emph{ad hoc} constructs and dialects are less frequently necessary. %thereby reducing the need for language dialects and revisions.

%Verse features a module system taken directly from SML. Unlike SML, the Verse core language is split into a \emph{typed external language} (EL) specified by {type-directed translation} to a minimal \emph{typed internal language} (IL). 
\begin{enumerate}
\item \textbf{Typed syntax macros} (TSMs), introduced in Sec. \ref{sec:tsms}, reduce the need to primitively build in derived concrete syntactic forms specific to library constructs (e.g. list syntax as in SML or XML syntax as in Scala and Ur/Web), by giving library providers static control over the parsing and expansion of delimited segments of textual syntax (at a specified type or parameterized family of types). 
\item \textbf{Type-specific languages} (TSLs), introduced in Sec. \ref{sec:tsls}, further reduce syntactic cost by allowing library providers to associate a TSM with a type declaration and then rely on a local type inference scheme to invoke that TSM implicitly.
%\item \textbf{Metamodules}, introduced in Sec. \ref{sec:metamodules}, reduce the need to primitively build in the type structure of constructs like records (and variants thereof),  labeled sums and other interesting constructs that we will introduce later by giving library providers programmatic ``hooks'' directly into the semantics, which are specified as a \emph{type-directed translation semantics} targeting a small \emph{typed internal language} (introduced in Sec. \ref{sec:verse}). %For example, a library provider can implement the type structure of records with a metamodule that:
%\begin{enumerate}
%\item introduces a type constructor, \lstinline{record}, parameterized by finite mappings from labels to types, and defines, programmatically, a translation to unary and binary product types (which are built in to the internal language); and 
%\item introduces operators used to work with records, minimally record introduction and elimination (but perhaps also various functional update operators), and directly implements the logic governing their typechecking and translation to the IL (which builds in only nullary and binary products). 
%\end{enumerate}
%We will see direct analogies between ML-style modules (which our mechanisms also support) and metamodules later.
\end{enumerate} 
As vehicles for this work, we plan to formally specify a series of small typed lambda calculi that capture each of the novel primitives that we introduce ``minimally''. For the sake of examples, we will also describe (but not formally specify) a ``full-scale'' functional language called Verse.\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work that we will describe, because Wyvern is a group effort evolving independently in some important ways.} Verse is based semantically on Standard ML, differing mainly in that it uses a local type inference scheme \cite{Pierce:2000:LTI:345099.345100} (like, for example, Scala) for reasons that we will return to in Sec. \ref{sec:tsls}. The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal specification of Verse is both to emphasize that the primitives we introduce can be considered for inclusion in a variety of language designs, and to avoid distracting the reader with specifications for ``orthogonal'' primitives that are already well-understood in the literature. %We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble. %The purpose of the work being proposed is to serve as a reference for those interested in the new constructs we introduce, not to serve as a language specification. 
%We will give a brief overview of these languages are organized in Sec. \ref{sec:verse}.

The primitives we propose perform \emph{static code generation} (also sometimes called \emph{static} or \emph{compile-time metaprogramming}), meaning that the relevant rules in the static semantics of the language call for the evaluation of \emph{static functions} that generate static representations of terms. Static functions are functions written in a restricted subset of the language (we will discuss the design space of restrictions in Sec. \ref{sec:tsms-limitations}). %Library providers write these static functions using the Verse \emph{static language} (SL).  
The design we are proposing also has conceptual roots in earlier work on \emph{active libraries}, which similarly envisioned using compile-time computation to give library providers more control over aspects of the language (but did not take a type theoretic approach) \cite{active-libraries-thesis}. %Maintaining a separation between the static (or ``compile-time'') phase and the dynamic (or ``run-time'') phase is an important facet of Verse's design. % static code generation. %We will  also introduce a simple variant of each of these primitives that leverages Verse's support for local type inference to further reduce syntactic cost in certain common situations.

The main challenge in the design of these primitives will come in ensuring that they are metatheoretically well-behaved. If we are not careful, many of the problems  that arise when combining language dialects, discussed earlier, could simply shift into the semantics of these primitives.\footnote{This is why languages  like Verse are often called ``extensible languages'', though this is somewhat of a misnomer. The defining characteristic of an extensible language is that it \emph{doesn't} need to be extended in situations where other languages would need to be extended. We will avoid this somewhat confusing terminology.} Our main technical contributions will be in rigorously showing how to address these problems in a principled manner. In particular, syntactic conflicts will be impossible by construction and the semantics will validate code statically generated by TSMs and TSLs to maintain:
\begin{itemize}
\item a \emph{hygienic type discipline}, meaning that the language is type safe, that one can reason about the type of a well-typed expression without examining its expansion, and that the expansion does not make any assumptions about the names of variables in the surrounding context; and
\item \emph{modular reasoning principles}, meaning that library providers will have the ability to reason about the constructs that they have defined in isolation, and clients will be able to use them safely in any combination, without the possibility of conflict.\footnote{This is not quite true -- simple naming conflicts can arise. We will tacitly assume that they are being avoided extrinsically, e.g. by using a URI-based naming scheme as in the Java ecosystem.} 
\end{itemize}
%We will make these notions completely precise as we continue.

\subsection{Thesis Statement}
In summary, we propose a thesis defending the following statement:
\begin{quote}
A functional programming language can give library providers the ability to %meta\-pro\-gram\-matic\-ally 
express new syntactic expansions while maintaining a hygienic type discipline and modular reasoning principles. %These  primitives are  expressive enough to subsume the need for a variety of primitives that are, or would need to be, built in to comparable contemporary languages.
\end{quote}

>>>>

%Well-designed general-purpose programming languages give library providers the power to  modularly express a wide variety of useful constructs using a comparatively small set of primitives. %New primitives, and thus new dialects of these languages, should arise very rarely. 
%New primitives (and new \emph{dialects} of the language). %Extending a general-pandurpose language with new primitives, i.e.  forming an extended \emph{dialect} of the language, should, given a sufficiently expressive such language, be exceedingly rare. %Indeed, a diversity of dialects around a language can be taken as evidence against claims of  generality.%, this can be taken as evidence against claims of its generality. %Not all useful abstractions can be  realized, when considered comprehensively, as libraries.% The problem of achieving both simplicity and generality simultaneously with a single system remains unsolved. 
%Many languages claim to achieve this ideal. 
%For example, 
%If we measure the generality of a language by how frequently  dialects of the language emerge, 
%all major contemporary languages seem to leave room for improvement. 
%General-purpose programming languages are a dime a. 
%Among the chief goals in the design of a general-purpose programming languages must achieve \emph{stability}: they  very rarely needs to be extended or forked into dialects by the user community surrounding it. Instead, its users are able to modularly  express a wide variety of useful constructs in libraries, using the comparatively small set of primitives that the language builds in. 
Functional programming languages come in many sizes. The smallest functional languages -- often referred to as ``{lambda calculi}'' -- allow researchers and language designers to study the mathematical properties of language primitives of interest in isolation. These ``small-scale'' languages inform the design of ``full-scale'' 
%\footnote{Throughout this work, words and phrases that should be read as having an intuitive or informal meaning, rather than a strict mathematical meaning, will be introduced with quotation marks.} 
 languages, which combine several such primitives and also define various derived syntactic forms (colloquially, ``syntactic sugar'') that decrease the syntactic cost of selected idioms. 
For example, Standard ML (SML) \cite{mthm97-for-dart,harper1997programming}, OCaml \cite{ocaml-manual} and Haskell \cite{jones2003haskell}   build in 
%record types, generalizing the nullary and binary product types that suffice in simpler calculi, because labeled components are cognitively useful to human programmers. Similarly, these languages all build in 
derived forms that decrease the syntactic cost of working with lists. For example, in these languages, the form \lstinline{[1, 2, 3, 4, 5]} desugars to: 
\begin{lstlisting}[numbers=none]
Cons(1, Cons(2, Cons(3, Cons(4, Cons(5, Nil)))))
\end{lstlisting}

The hope amongst many language designers is that a limited number of derived forms like these will suffice to produce a ``general-purpose'' programming language, i.e. one that satisfies programmers working in a wide variety of application domains. Unfortunately, a stable language design that fully achieves this ideal has yet to emerge, as evidenced by the diverse array of \emph{syntactic dialects} -- dialects that introduce only new derived forms -- that continue to proliferate around all major contemporary languages. For example, Ur/Web is a syntactic dialect of Ur (an ML-like full-scale language \cite{conf/pldi/Chlipala10}) that builds in derived forms for SQL queries, HTML elements and other datatypes used in the domain of web programming \cite{conf/popl/Chlipala15}. %Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined. 
We will consider another example -- regular expression patterns -- in Sec. \ref{sec:examples}. 
Tools like Camlp4 \cite{ocaml-manual}, Sugar* \cite{erdweg2011sugarj,erdweg2013framework} and Racket's preprocessor \cite{Flatt:2012:CLR:2063176.2063195}, which we will discuss in Sec. \ref{sec:syntax-existing-approaches}, have decreased the engineering costs of constructing syntactic dialects, further contributing to their proliferation.

\subsection{Dialects Considered Harmful}
% express record types as syntactic sugar over the simply-typed lambda calculus with  binary product types.\footnote{Pairs can of course be expressed as syntactic sugar atop records, though one could argue that using binary products as the more primitive concept is simpler.} The static semantics need to be extended with new type and term operators. However, the simplest way to express the dynamic semantics of the newly introduced term operators is by translation to nested binary products, so we can leave the operational semantics alone. \todo{fill this out} %For example, there are dozens of constructs that go by the name of ``records'' in various languages, each defined by a slightly different collection of primitive operations. \todo{examples} %, encouraged  historically  by the availability of tools like compiler generators and,  more recently, language workbenches \cite{workbenches} and DSL frameworks \cite{dsl}. Unfortunately, taking this approach makes it substantially more difficult for clients to import high-level abstractions orthogonally. 
% test 
Some  view this proliferation of dialects as harmless or even as desirable, arguing that programmers can simply choose the right dialect for the job at hand \cite{journals/stp/Ward94}. However, this ``dialect-oriented'' approach is, in an important sense, anti-modular: programmers cannot always ``combine'' different dialects if they want to use the primitives that they feature together in a single program. For example, a programmer might have access to a syntactic dialect featuring HTML syntax and one featuring regular expression syntax, but it may not be straightforward to construct a dialect featuring both.

In some cases, coming up with a combined dialect is difficult because the dialects are specified using different formalisms. In other cases, a common formalism has been used, but it does not operationalize the notion of dialect combination (e.g. Racket's preprocessor \cite{Flatt:2012:CLR:2063176.2063195}). But even if we restrict our interest to dialects specified using a common formalism that does operationalize some notion of dialect combination, there may still be a problem: the formalism may not guarantee that the combined dialect will conserve important properties that can be established about the dialects in isolation. %In other words, any putative ``combined language'' must formally be considered a  distinct system for which one must derive essentially all metatheorems of interest anew, guided only informally by those derived for the dialects individually. %There is no well-defined mechanism for constructing such a ``combined language'' in general. 
For example, consider two syntactic dialects specified using Camlp4, one specifying derived syntax for finite mappings, the other specifying overlapping syntax for \emph{ordered} finite mappings. Though each dialect has an unambiguous (a.k.a. deterministic) grammar, when these grammars are na\"ively  combined,  ambiguities will arise. We are aware of only one formalism that guarantees that determinism is conserved when syntactic dialects are combined \cite{conf/pldi/SchwerdfegerW09}, but it has limited expressive power, as we will discuss in Sec. \ref{sec:direct-syntax-extension}.
%It is thus infeasible to simply allow different contributors to a software system to choose their own favorite dialect for each component they are responsible for. 
%It it clear that dialects are better rhetorical devices than practical engineering artifacts.

%Due to this paucity of modular reasoning principles, the ``dialect-oriented'' approach is problematic for software development ``in the large''. %Large software projects and software ecosystems must pick a single language that does provide powerful modular reasoning principles and, to benefit from them, stay inside it.

\subsection{Large Languages Considered Harmful}
Dialects sometimes have a less direct influence on large-scale software development: they can help convince the designers in control of comparatively popular languages, like OCaml and Scala, to include some variant of the primitives that they feature into backwards-compatible language revisions. %These decisions are increasingly influenced by community processes, e.g. the Scala Improvement Process.  %This approach concentrates power as well as responsibility over maintaining metatheoretic guarantees in the hands of a small group of language designers, though increasingly influenced by various community processes (e.g. the Scala Improvement Process). 
%Dialects thus serve the role of rhetorical vehicles for new ideas, rather than direct artifacts. 
%Over time, accepting such extensions has caused these languages to balloon in size. 
This \emph{ad hoc} approach is not sustainable, for three main reasons. First, we have gathered empirical data that suggests that there are simply too  many potentially useful such primitives \cite{TSLs}. Many of these are only relevant in relatively narrow application domains. It is unreasonable to expect language designers to be able to evaluate all of these use cases in a timely and informed manner. Second, primitives introduced earlier in a language's lifespan can end up monopolizing finite ``syntactic resources'', forcing subsequent primitives to use ever more esoteric forms. And third, primitives that prove after some time to be flawed in some way cannot be removed or modified without breaking backwards compatibility. For all these reasons, language designers are justifiably reticent to add new primitives to major programming languages.%Because there is often no empirical data about how useful a construct is in practice until it is available in a major language, decisions about which constructs to include are often informed only by intuition (and are thus)
%Recalling the words of  Reynolds, which are clearly as relevant today as they were almost half a century ago \cite{Reynolds70}: %This approach is antithetical to the ideal of a truly \emph{general-purpose language} described at the beginning of this section.
%\newpage

\subsection{Toward More General Primitives}
If, as we've argued, language designers should strive both to avoid dialect formation and to keep general-purpose languages small, stable and free of \emph{ad hoc}  primitives, this leaves two possible paths forward. One, exemplified (arguably) by SML, is to simply eschew ``niche'' embellishments and settle on the existing design, which might be considered to sit at a ``sweet spot'' in the overall language design space (accepting that in some circumstances, this leads to  high syntactic cost). 
The other path forward is to search for a small number of highly general primitives that allow us degrade many of the constructs that are built primitively into languages and their dialects today instead to modular library constructs. 
Encouragingly, primitives of this sort do occasionally arise. For example, a recent revision of OCaml added support for  generalized algebraic data types (GADTs), based on research on guarded recursive datatype constructors \cite{XiCheChe03}. Using GADTs, OCaml was able to move some of the \emph{ad hoc} machinery for typechecking operations that use format strings, like \li{sprintf}, out of the language and into a library. Syntactic machinery related to \li{sprintf}, however, remains  built in.

\section{Proposed Contributions}
%Our broad aim in the work being proposed is to introduce primitive language mechanisms that give library providers the ability to  express new syntactic expansions as well as new types and operators in a safe and modularly composable manner. 
Our aim in the work being proposed is to introduce primitive language constructs that take further steps down the second path just described by reducing the need for syntactic dialects. In particular, we plan to introduce the following primitives:
% By supporting the primitives that we introduce, 1) VerseML will be smaller than comparable languages like ML and Scala, and 2) dialect formation will be less frequently necessary. In other words, these primitives reduce the need for many others:%This eliminates the needs to build in fewer \emph{ad hoc} constructs and dialects are less frequently necessary. %thereby reducing the need for language dialects and revisions.

%VerseML features a module system taken directly from SML. Unlike SML, the VerseML core language is split into a \emph{typed external language} (EL) specified by {type-directed translation} to a minimal \emph{typed internal language} (IL). 
\begin{enumerate}
\item \textbf{Typed syntax macros} (TSMs), introduced in Sec. \ref{sec:tsms}, reduce the need to primitively build in derived concrete syntactic forms specific to library constructs (e.g. list syntax as in SML or XML syntax as in Scala and Ur/Web), by giving library providers static control over the parsing and expansion of delimited segments of textual syntax (at a specified type or parameterized family of types). 
\item \textbf{Type-specific languages} (TSLs), introduced in Sec. \ref{sec:tsls}, further reduce syntactic cost by allowing library providers to associate a TSM with a type declaration and then rely on a local type inference scheme to invoke that TSM implicitly.
%\item \textbf{Metamodules}, introduced in Sec. \ref{sec:metamodules}, reduce the need to primitively build in the type structure of constructs like records (and variants thereof),  labeled sums and other interesting constructs that we will introduce later by giving library providers programmatic ``hooks'' directly into the semantics, which are specified as a \emph{type-directed translation semantics} targeting a small \emph{typed internal language} (introduced in Sec. \ref{sec:VerseML}). %For example, a library provider can implement the type structure of records with a metamodule that:
%\begin{enumerate}
%\item introduces a type constructor, \lstinline{record}, parameterized by finite mappings from labels to types, and defines, programmatically, a translation to unary and binary product types (which are built in to the internal language); and 
%\item introduces operators used to work with records, minimally record introduction and elimination (but perhaps also various functional update operators), and directly implements the logic governing their typechecking and translation to the IL (which builds in only nullary and binary products). 
%\end{enumerate}
%We will see direct analogies between ML-style modules (which our mechanisms also support) and metamodules later.
\end{enumerate} 
As vehicles for this work, we plan to formally specify a series of small typed lambda calculi that capture each of the novel primitives that we introduce ``minimally''. For the sake of examples, we will also describe (but not formally specify) a ``full-scale'' functional language called VerseML.\footnote{We distinguish VerseML from Wyvern, which is the language referred to in prior publications about some of the work that we will describe, because Wyvern is a group effort evolving independently in some important ways.} VerseML is, as its name suggests, a dialect of ML. It diverges from other dialects of ML that have a similar underlying type structure, like Standard ML and OCaml, in that it uses a local type inference scheme \cite{Pierce:2000:LTI:345099.345100} (like, for example, Scala \cite{OdeZenZen01}) for reasons that have to do with the mechanisms described in Sec. \ref{sec:tsls}. The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal specification of VerseML in this work is both to emphasize that the primitives we introduce can be considered for inclusion in a variety of language designs (not exclusively dialects of ML), and to avoid distracting the reader with specifications for ``orthogonal'' primitives that are already well-understood in the literature. %We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble. %The purpose of the work being proposed is to serve as a reference for those interested in the new constructs we introduce, not to serve as a language specification. 
%We will give a brief overview of these languages are organized in Sec. \ref{sec:VerseML}.

The primitives we propose perform \emph{static code generation} (also sometimes called \emph{static} or \emph{compile-time metaprogramming}), meaning that the relevant rules in the static semantics of the language call for the evaluation of \emph{static functions} that generate static representations of terms. Static functions are functions written in a restricted subset of the language (we will discuss the design space of restrictions in Sec. \ref{sec:tsms-limitations}). %Library providers write these static functions using the VerseML \emph{static language} (SL).  
The design we are proposing also has conceptual roots in earlier work on \emph{active libraries}, which similarly envisioned using compile-time computation to give library providers more control over aspects of the language (but did not take a type theoretic approach) \cite{active-libraries-thesis}. %Maintaining a separation between the static (or ``compile-time'') phase and the dynamic (or ``run-time'') phase is an important facet of VerseML's design. % static code generation. %We will  also introduce a simple variant of each of these primitives that leverages VerseML's support for local type inference to further reduce syntactic cost in certain common situations.

The main challenge in the design of the primitives we introduce will come in ensuring that they are metatheoretically well-behaved. If we are not careful, many of the problems  that arise when combining language dialects, discussed earlier, could simply shift into the semantics of these primitives.\footnote{This is why languages  like VerseML are often called ``extensible languages'', though this is somewhat of a misnomer. The defining characteristic of an extensible language is that it \emph{doesn't} need to be extended in situations where other languages would need to be extended. We will avoid this somewhat confusing terminology.} Our main technical contributions will be to rigorously address these problems in a principled manner. In particular, we will maintain:
\begin{itemize}
\item a \emph{type discipline}, meaning that the language is type safe, and that programmers reading a well-typed expression can determine its type  without examining its expansion; 
\item a \emph{hygiene discipline} (a.k.a. a \emph{binding discipline}), meaning that the expansion logic does not make any assumptions about the names of variables in the context surrounding the expansion, nor does it introduce ``hidden bindings'' into subexpressions; and 
\item \emph{modular reasoning principles}, meaning that library providers will have the ability to reason about the constructs that they have defined in isolation, and clients will be able to use them safely in any combination, without the possibility of conflict.\footnote{This is not quite true -- simple naming conflicts can arise. We will tacitly assume that they are being avoided extrinsically, e.g. by using a URI-based naming scheme as in the Java ecosystem.} 
\end{itemize}
%We will make these notions completely precise as we continue.

\subsection{Thesis Statement}
In summary, we propose a thesis defending the following statement:
\begin{quote}
A functional programming language can give library providers the ability to %meta\-pro\-gram\-matic\-ally 
express new syntactic expansions while maintaining a type discipline, a hygiene discipline and modular reasoning principles.%These  primitives are  expressive enough to subsume the need for a variety of primitives that are, or would need to be, built in to comparable contemporary languages.
\end{quote}

\subsection{Disclaimers}
Before we continue, it may be useful to explicitly acknowledge that completely eliminating the need for dialects would indeed be asking for too much: certain design decisions are fundamentally incompatible with others or require coordination across a language design. We aim only to decrease the need for syntactic dialects in this work. % out a larger design space within a single language, VerseML.%a subset of constructs that can be specified by a semantics of a certain ``shape'' specified by VerseML (we will make this more specific later). %There is nothing ``universal'' about VerseML.

It may also be useful to explicitly acknowledge that library providers could leverage the primitives we introduce   to define constructs that are in ``poor taste''. We  expect that in practice, VerseML will come with a standard library defining a carefully curated collection of standard constructs, as well as guidelines for advanced users regarding when it would be sensible to use the mechanisms we introduce (following the example of languages that support operator overloading or type classes \cite{Hall:1996:TCH:227699.227700}, which also have the potential for ``abuse''). For most programmers, using VerseML should not be substantially different from using a language like ML or one of its dialects.%The vast majority of programmers should not use the primitives that we introduce directly.

Finally, VerseML intentionally is not a dependently-typed language like Coq, Agda or Idris, because these languages do not maintain a phase separation between ``compile-time'' and ``run-time.'' This phase separation is useful for programming tasks (where one would like to be able to discover errors before running a program, particularly programs that may have an effect) but less so for theorem proving tasks (where it is mainly the fact that a pure expression is well-typed that is of interest, by the propositions-as-types principle). VerseML is designed to be used for programming tasks where SML, OCaml, Haskell or Scala would be used today, not for advanced theorem proving tasks. That said, we conjecture that the primitives we describe could be added to languages like Gallina (the ``external language'' of the Coq proof assistant  \cite{Coq:manual}) with  modifications, but do not plan to pursue this line of research in this dissertation. %Our interest is in making sure that the standard library is no more privileged than any other library. % Our interest is not in creating a universal language, only one that is as expressive as reasonably possible. %only in a subset of all constructs that can be specified in a mutually orthogonal manner by a semantics of a certain prototypic ``shape''. We will make this more specific later.

% \section{Language Overview}\label{sec:VerseML}
% Let us begin with a brief overview of how VerseML and the minimal calculi that we will develop are organized and specified at a high level. We will assume that readers are familiar with the early chapters of a textbook like \emph{TAPL} \cite{tapl} or \emph{PFPL} \cite{pfpl}.%VerseML consists of a \emph{module language} atop a \emph{core language}.
